A Bidirected String Graph Model for Genome Assembly

### 2012-12-12 presentation/paper.tex
paper: updates

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
\usepackage{threeparttable}
%\usepackage{hyperref}
%\usepackage{amsthm}

@@

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
length of any individual read, which makes the genome more difficult to assemble
accurately.  However, real genomes almost always contain repeats of substantial
length, so assemblers must be able to handle repeats.

@@

Genome assembly is superficially similar to the shortest common superstring
problem, which is to find the shortest possible string that contains every
string in a given set as substrings.  Genome assembly also requires finding a
longer string that, roughly, contains every string in  a set of substrings.
However, the actual genome may be more than one string, and the genome may also
be longer than the shortest possible genome from which the reads could have been
sampled, due to repeats.  In addition,  genomes are double-stranded, and there
may be other complications with the reads, such as sequencing error.  Therefore,
genome assembly is not the same as the shortest common superstring problem, and
in fact seems to be harder, despite the fact that the shortest common
superstring problem is already NP-complete\cite{Turner1989}.

@@

This paper explores the genome assembly problem through an algorithm published
in 2005 by Eugene Myers that is based on the idea of the fragment string
assembly graph\cite{Myers2005}.  The algorithm provides a good overview of the
techniques that are used in genome assembly, especially since it draws on ideas
used in previous assemblers, such as the Celera assembler, which was used to
assemble the human genome for the first time\cite{Venter2001}.  In addition, the
algorithm shares some ideas with de Bruijn graph-based assemblers, such as
EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
assembly model.  In fact, the algorithm is called the {\em fragment} string
graph algorithm because the graph is built directly from the sequenced fragments
(or reads), rather than subsequences ($k$-mers) taken from them.

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from condition (4) is
{\em not} a reverse-complement overlap, as it can equivalently be stated as the
last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
may arise from this definition.

@@

Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
overlapped regions of the two reads do not match exactly (in either a forward or
reverse-complement manner), but instead differ by a number of base-pair
substitutions, insertions, and/or deletions up to a predetermined upper bound,
such as 5\%.  However, inexact overlaps will not be considered further in this
paper.

@@

The naive way to compute overlaps is to consider every pair of reads in the
input, including only distinct pairs but also including the special cases of
each read paired with itself, and search for each of the types of overlaps
between the two reads in each pair.  If there are $N$ reads in the input, this
algorithm will need to consider $\frac{N^2}{2} + N$ pairs of reads. The time
spent on each read pair will depend several factors, including the minimum
overlap length $L$ and the read lengths.  But even if we could somehow spend
constant time on each read pair, this naive algorithm would be prohibitively
slow because the running time rises quadraticly with respect to the number of
reads, and there can easily be millions of reads in the input.

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled together when the overlaps
are used as evidence.  To do this, define a graph where each vertex corresponds
to a read\footnote{Actually, more than one read may map to the same vertex, as
explained in \Subsection{subsec:removing_contained_reads}.} and each edge
corresponds to an overlap.  Two vertices are connected by an edge if and only if
the corresponding reads share an overlap.  In addition, each edge is labeled
with an appropriate DNA sequence, the details of which will be explained later.

@@

\begin{figure}
	\begin{center}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[->,style=very thick] (f) edge (g);
				\draw[->,style=very thick] (g) edge (h);
				\draw[->,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible directed \\representation of the \\string graph.}
			\label{subfig:possible_directed}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[-,style=very thick] (f) edge (g);
				\draw[-,style=very thick] (g) edge (h);
				\draw[-,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible undirected \\representation of the \\string graph.}
			\label{subfig:possible_undirected}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[>->,style=very thick] (f) edge (g);
				\draw[>-<,style=very thick] (g) edge (h);
				\draw[>-<,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible bidirected \\representation of the \\string graph.}
			\label{subfig:possible_bidirected}
		\end{subfigure}
		\caption{Three possibilities for constructing the string graph.  The
		bidirected construction works the best in terms of modeling the
		underlying problem.}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.00, 0) {2};
			\node (3) at (4.00, 0) {3};
			\node (4) at (6.00, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
			\draw[color=red,thick] (2.00, 0) circle (0.8);
			\draw[color=red,thick] (4.00, 0) circle (0.8);
		\end{tikzpicture}
	} \end{center}
	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
	walks in the bidirected graph above.  (Check circled pairs of heads to
	verify this.)}
	\label{fig:bidiwalk_example}
\end{figure}

@@

The concept of a walk in a bidirected graph maps quite nicely onto the genome
assembly problem because a walk through the bidirected string graph constructed
from a set of reads and their overlaps models a way in which the reads can be
assembled together consistently\cite{Myers2005}.  In such a walk, each read may
be used in forward or reverse-complement orientation, depending on the
direction, relative to the walk, of the adjacent arrow heads when the
corresponding vertex is passed through.

@@

\begin{figure}
	\begin{center}
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.0, 0) {2};
			\node (3) at (4.0, 0) {3};
			\node (4) at (6.0, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} }

@@

\begin{figure}
	\begin{center}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
				\draw[>->,style=very thick] (f) edge (g);
				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.00, -0.1) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
				\draw[->] (0.5, 0.5) ..
					controls (0.7, 0.7) and (1.3, 0.7) ..
					(1.5, 0.5);
				\draw[->] (1.5, -0.5) ..
					controls (1.3, -0.7) and (0.7, -0.7) ..
					(0.5, -0.5);
			\end{tikzpicture}
			\caption{A bidirected string graph fragment}
			\label{subfig:bidiedge_interp_a}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
				\draw[->,very thick] (-0.2, 0) node[anchor=east] {$f.B$}
						-> (4, 0) node[anchor=west] {$f.E$};
				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
						-> (6.29, -0.9) node[anchor=west] {$g.E$};

@@

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (fE) at (0, 0) {f.E};
				\node[circle,fill=blue!20,draw=black] (gE) at (2.0, 0) {g.E};

				\node[circle,fill=blue!20,draw=black] (fB) at (0, -1) {f.B};
				\node[circle,fill=blue!20,draw=black] (gB) at (2.0, -1) {g.B};

				\draw[->,style=very thick] (fE) edge (gE);
				\draw[->,style=very thick] (gB) edge (fB);
				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.10, -1.01) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
			\end{tikzpicture}
		\end{center}
		\caption{The corresponding directed string graph fragment}
		\label{subfig:bidiedge_interp_c}
	\end{subfigure}

@@

Earlier, the definition of a \KeyTerm{contained read} was provided.  Contained
reads need not be included in the graph because there will always be a walk that
spells the sequence of each contained read.  In other words, contained reads
provide redundant information because the sequence of any given contained read
is already spelled out by an uncontained read.

Therefore, prior to mapping the reads to vertices and overlaps to edges, we can
examine the overlaps to identify contained reads, and include in the initial
construction of the bidirected string graph neither the contained reads nor any
overlaps involving a contained read.

This implies that a vertex in the bidirected string graph may actually
correspond to more than one read.

@@

Building the graph is implemented by either the program
\ProgramName{build-bidirected-string-graph} or the program
\ProgramName{build-directed-string-graph}, the latter of which builds the
directed equivalent to the bidirected graph described in
\Subsection{subsec:directed}.  Both programs take as input two files: the set of
uncontained reads and their overlaps, and produce as output the bidirected or
directed graph, serialized to a file.  As should be expected from what was
described earlier, \ProgramName{build-bidirected-string-graph} will add one
vertex for each read and add a bidirected edge for each overlap, while
\ProgramName{build-directed-string-graph} will add two vertices for each read
and add two directed edges for each overlap.

The string graphs are stored in a binary format.  An additional program,
\ProgramName{print-string-graph}, was implemented that can print a string graph
in plain-text format, print a string graph in the DOT language for input to the
graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print some
other information about a string graph.

@@

A program \ProgramName{extract-edge-seqs} was implemented to extract the
sequences that the edges are labeled with, as this can be done to produce a
preliminary assembly (especially after the graph simplifications that are about
to be described).

@@

@@

Before trying to find walks through the string graph, it is very helpful to
simplify the graph is much as possible.  Very commonly, given three adjacent
reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
because we can, equivalently, walk $f \to g \to h$ rather than directly walking
from $f \to h$.  \Figure{fig:transitive} shows an example of this.

\begin{figure}
	\begin{center}
		\begin{tikzpicture}
			\draw[->,style=very thick] (0,0) node[anchor=east] {f} -- (2,0);
			\draw[->,style=very thick] (0.6,0.25) node[anchor=east] {g} -- (2.6,0.25);
			\draw[->,style=very thick] (1.2,0.5) node[anchor=east] {h} -- (3.2,0.5);
		\end{tikzpicture}
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (f) at (0, 0) {f};
			\node (g) at (1.7, 0) {g};
			\node (h) at (3.4, 0) {h};
			\draw[>->,style=very thick] (f) edge (g);
			\draw[>->,style=very thick] (g) edge (h);
			\draw[>->,style=very thick,bend left] (f) edge (h);
		\end{tikzpicture}
	\end{center}
	\caption{Three overlapping reads $f$, $g$, and $h$, and the corresponding
	bidirected graph.  The edge between $f$ and $h$ is transitive and is removed
	by transitive reduction.}
	\label{fig:transitive}
\end{figure}

@@

\Algorithm{alg:transitive_reduction} provides pseudocode for a transitive
reduction algorithm similar to that described in \cite{Myers2005}.  Essentially,
the algorithm is to through each vertex $v$ and examine vertices up to 2 edges
away from $v$ to identify all transitive edges leaving $v$.  These edges are
then marked for removal, and they are removed after all vertices in the graph
have been considered.  Some optimizations are possible; for example, the lengths
of the sequences the edges are labeled with can be taken into account to allow
terminating the search of $v$'s neighborhood before all possible walks of length
2 starting at $v$ have been considered.

@@

\begin{figure}
	\begin{center}
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node[fill=green!25] (4) at (3.25, 0) {4};
			\node (5) at (5.0, 0) {5};
			\node (6) at (6.5, 1) {6};
			\node (7) at (6.5, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (4);
			\draw[>->,very thick] (4) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}

@@

		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node (5) at (3.25, 0) {5};
			\node (6) at (4.75, 1) {6};
			\node (7) at (4.75, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}
	\end{center}
	\caption{Vertex 4 an \KeyTerm{inner vertex} that is part of an unbranched
	path that can be collapsed.  Edge labels are not shown.}
	\label{fig:collapse_unbranched_paths}
\end{figure}

@@

A completely random genome is very unlikely to have repeats of significant
length compared to the minimum overlap length.  In this case, transitive
reduction is expected to leave the entire graph as a single unbranched path,
provided the original genome was a single, non-circular string and every base
pair of it was included in at least one read.  Then, collapsing unbranched paths
will collapse the entire bidirected graph into two vertices connected by a
single edge.  Finding a walk through the edges of this graph is trivial and will
produce an assembly of the entire genome.  \Figure{fig:random_example} shows the
bidirected string graph that might result from such a genome and data.

If we still assume a genome with no repeats of length greater than the overlap
length, but relax the assumption that the genome is a single, non-circular
string from which every base pair was included in a read, the graph will instead
reduce and collapse to a set of components.  Components corresponding to
non-circular sequence will contain two vertices with an edge between, while
components corresponding to circular sequence will contain one vertex with an
edge connecting it to itself (a loop).  In this case (and as will be elaborated
on later), finding a walk through the bidirected string graph must be
generalized to allow walking each component separately; but again for this
particular case of a random genome, each such walk is still trivial, and the
assembly will be optimal as long as the assembler does not decide to traverse
any of the loops more than one time (this will also be elaborated on more
later), since there are no repeats of substantial length in a hypothetical
random genome.

>>>>

\documentclass[10pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
\usepackage{threeparttable}
%\usepackage{hyperref}
%\usepackage{amsthm}

%\newcommand{\nodecolor}{blue!20}
\newcommand{\nodecolor}{black!0}
%\newcommand{\specialnodecolor}{green!25}
\newcommand{\specialnodecolor}{black!10}

@@

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  The number of times a
subsequence occurs in the genome is referred to as its \KeyTerm{copy number}.
Repeats may be longer than the length of any individual read, which makes the
genome more difficult to assemble accurately.  However, real genomes almost
always contain repeats of substantial length, so assemblers must be able to
handle repeats.

The output of an assembler may be a list of \KeyTerm{contigs}, which are
continuous sequences of DNA that were assembled and therefore are believed to
exist in the original genome.  Each contig should be much longer than the
original reads.  Ideally, if the original genome contains $N$ separate strings,
there should be $N$ contigs, each of which is one of the genomic strings.
However, in general there will be more than $N$ contigs because it is rarely
possible to perfectly assemble the genome.

@@

Genome assembly is superficially similar to the shortest common superstring
problem, which is to find the shortest possible string that contains every
string in a given set as substrings.  Genome assembly also requires finding a
longer string that, roughly, contains every string in  a set of substrings.
However, the actual genome may be more than one string, and the genome may also
be longer than the shortest possible genome from which the reads could have been
sampled, due to repeats.  In addition,  genomes are double-stranded, and there
may be other complications with the reads, such as sequencing error.  Therefore,
genome assembly is not the same as the shortest common superstring problem, and
in some ways seems to be harder, despite the fact that the shortest common
superstring problem is already NP-complete\cite{Turner1989}.

@@

This paper explores the genome assembly problem through an algorithm published
in 2005 by Eugene Myers that is based on the idea of the fragment string
assembly graph\cite{Myers2005}.  The algorithm provides a good overview of the
techniques that are used in genome assembly, especially since it draws on ideas
used in previous assemblers, such as the Celera assembler, which was used to
assemble the human genome for the first time\cite{Venter2001}.  In addition, the
algorithm shares some ideas with de Bruijn graph-based assemblers, such as
EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
assembly model.  In fact, the algorithm is called the {\em fragment} string
graph algorithm because the graph is built directly from the sequenced fragments
(reads), rather than subsequences ($k$-mers) taken from them.

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from condition (4) is
{\em not} a reverse-complement overlap, as it can equivalently be stated as the
last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$ bp of $f$.
\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
may arise from this definition.

@@

Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
the overlapped regions of the two reads do not match exactly (in either a
forward or reverse-complement manner), but instead differ by a number of
base-pair substitutions, insertions, and/or deletions up to a predetermined
upper bound, such as 5\%.  However, inexact overlaps will not be considered
further in this paper.

@@

The naive way to compute overlaps is to consider every pair of reads in the
input, including only distinct pairs but also including the special cases of
each read paired with itself, and search for each of the types of overlaps
between the two reads in each pair.  If there are $N$ reads in the input, this
algorithm will need to consider $\frac{N^2}{2} + N$ pairs of reads. The time
spent on each read pair will depend several factors, including the minimum
overlap length $L$ and the read lengths.  But even if we somehow could spend
constant time on each read pair, this naive algorithm would be prohibitively
slow because the running time rises quadraticly with respect to the number of
reads, and there can easily be millions of reads in the input.

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled when the overlaps are used as
evidence.  To do this, define a graph where each vertex corresponds to a
read\footnote{Actually, more than one read may map to the same vertex, as
explained in \Subsection{subsec:removing_contained_reads}.} and each edge
corresponds to an overlap.  Two vertices are connected by an edge if and only if
the corresponding reads share an overlap.  In addition, each edge is labeled
with an appropriate DNA sequence, the details of which will be explained later.

@@

\begin{figure}
	\begin{center}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[->,style=very thick] (f) edge (g);
				\draw[->,style=very thick] (g) edge (h);
				\draw[->,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible directed \\representation of the \\string graph}
			\label{subfig:possible_directed}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[-,style=very thick] (f) edge (g);
				\draw[-,style=very thick] (g) edge (h);
				\draw[-,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{A possible undirected \\representation of the \\string graph}
			\label{subfig:possible_undirected}
		\end{subfigure}
		\begin{subfigure}{0.32\textwidth}
			\begin{tikzpicture}[>=triangle 45]
				\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
				\node (f) at (0, 0) {f};
				\node (g) at (1.7, 0) {g};
				\node (h) at (3.4, 0) {h};
				\draw[>->,style=very thick] (f) edge (g);
				\draw[>-<,style=very thick] (g) edge (h);
				\draw[>-<,style=very thick,bend left] (f) edge (h);
			\end{tikzpicture}
			\caption{The bidirected \\representation of the \\string graph}
			\label{subfig:possible_bidirected}
		\end{subfigure}
		\caption{Three possibilities for constructing the string graph.  The
		bidirected construction works the best in terms of modeling the
		underlying problem.}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.00, 0) {2};
			\node (3) at (4.00, 0) {3};
			\node (4) at (6.00, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
			\draw[color=red,thick] (2.00, 0) circle (0.8);
			\draw[color=red,thick] (4.00, 0) circle (0.8);
		\end{tikzpicture}
	} \end{center}
	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
	walks in the bidirected graph above.  (Check circled pairs of heads to
	verify this.)}
	\label{fig:bidiwalk_example}
\end{figure}

@@

The concept of a walk in a bidirected graph maps quite nicely onto the genome
assembly problem because a walk through the bidirected string graph constructed
from a set of reads and their overlaps models a way in which the reads can be
assembled together consistently\cite{Myers2005}.  In such a walk, each read may
be used in either forward or reverse-complement orientation, depending on the
direction, relative to the walk, of the adjacent arrow heads when the
corresponding vertex is passed through.

@@

\begin{figure}
	\begin{center}
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.0, 0) {2};
			\node (3) at (4.0, 0) {3};
			\node (4) at (6.0, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} }

@@

\begin{figure}
	\begin{center}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=\nodecolor,draw=black] (f) at (0, 0) {f};
				\node[circle,fill=\nodecolor,draw=black] (g) at (2.0, 0) {g};
				\draw[>->,style=very thick] (f) edge (g);
				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.00, -0.1) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
				\draw[->] (0.5, 0.5) ..
					controls (0.7, 0.7) and (1.3, 0.7) ..
					(1.5, 0.5);
				\draw[->] (1.5, -0.5) ..
					controls (1.3, -0.7) and (0.7, -0.7) ..
					(0.5, -0.5);
			\end{tikzpicture}
			\caption{A bidirected string graph fragment}
			\label{subfig:bidiedge_interp_a}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
				\draw[->,very thick] (-0.2, 0) node[anchor=east] {$f.B$}
						-> (4, 0) node[anchor=west] {$f.E$};
				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
						-> (6.29, -0.9) node[anchor=west] {$g.E$};

@@

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=\nodecolor,draw=black] (fE) at (0, 0) {f.E};
				\node[circle,fill=\nodecolor,draw=black] (gE) at (2.5, 0) {g.E};

				\node[circle,fill=\nodecolor,draw=black] (fB) at (0, -1) {f.B};
				\node[circle,fill=\nodecolor,draw=black] (gB) at (2.5, -1) {g.B};

				\draw[->,style=very thick] (fE) edge (gE);
				\draw[->,style=very thick] (gB) edge (fB);
				\draw (1.2, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.2, -1.01) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
			\end{tikzpicture}
		\end{center}
		\caption{The corresponding directed string graph fragment}
		\label{subfig:bidiedge_interp_c}
	\end{subfigure}

@@

Earlier, the definition of a \KeyTerm{contained read} was provided.  Contained
reads need not be included in the graph because there will always be a walk that
spells the sequence of each contained read.
%In other words, contained reads
%provide redundant information because the sequence of any given contained read
%is already spelled out by an uncontained read.

Therefore, prior to mapping the reads to vertices and overlaps to edges, we can
examine the overlaps to identify contained reads, then include in the initial
construction of the bidirected string graph neither the contained reads nor any
overlaps involving a contained read.

This implies that a vertex in the bidirected string graph may actually
correspond to more than one read, which makes it meaningful to include a vertex
more than one time in a walk in the event of a genomic repeat.

@@

Building the graph is implemented by either the program
\ProgramName{build-bidirected-string-graph} or the program
\ProgramName{build-directed-string-graph}, the latter of which builds the
directed equivalent to the bidirected graph described in
\Subsection{subsec:directed}.  Both programs take as input two files: the set of
uncontained reads and their overlaps, and produce as output the bidirected or
directed graph, serialized to a file.
\ProgramName{build-bidirected-string-graph} will add one vertex for each read
and add a bidirected edge for each overlap, while
\ProgramName{build-directed-string-graph} will add two vertices for each read and
add two directed edges for each overlap.

The string graphs are stored in a binary format.  An additional program,
\ProgramName{print-string-graph}, was implemented that can print a string graph
in plain-text format, print a string graph in the DOT language for input to the
graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print other
information about a string graph.

@@

A program \ProgramName{extract-edge-seqs} was implemented to extract the
sequences that the edges are labeled with, as this can be done to produce a
preliminary assembly (especially after the graph simplifications described in
\Section{sec:simplify}).

@@

\label{sec:simplify}

@@

Before trying to find walks through the string graph, it is helpful to
simplify the graph as much as possible.  Very commonly, given three adjacent
reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
because we can, equivalently, walk $f \to g \to h$ rather than directly walking
$f \to h$.  \Figure{fig:transitive} shows an example of this.

\begin{figure}
	\begin{center}
		\begin{tikzpicture}
			\draw[->,style=very thick] (0,0) node[anchor=east] {f} -- (2,0);
			\draw[->,style=very thick] (0.6,0.25) node[anchor=east] {g} -- (2.6,0.25);
			\draw[->,style=very thick] (1.2,0.5) node[anchor=east] {h} -- (3.2,0.5);
		\end{tikzpicture}
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (f) at (0, 0) {f};
			\node (g) at (1.7, 0) {g};
			\node (h) at (3.4, 0) {h};
			\draw[>->,style=very thick] (f) edge (g);
			\draw[>->,style=very thick] (g) edge (h);
			\draw[>->,style=very thick,bend left] (f) edge (h);
		\end{tikzpicture}
	\end{center}
	\caption{Three overlapping reads $f$, $g$, and $h$, and the corresponding
	bidirected graph.  The edge between $f$ and $h$ is transitive and is removed
	by transitive reduction.}
	\label{fig:transitive}
\end{figure}

@@

\Algorithm{alg:transitive_reduction} provides pseudocode for a transitive
reduction algorithm similar to that described in \cite{Myers2005}.  Essentially,
the algorithm is to go through each vertex $v$ and examine vertices up to 2 edges
away from $v$ to identify all transitive edges leaving $v$.  These edges are
then marked for removal, and they are removed after all vertices in the graph
have been considered.  Some optimizations are possible; for example, the lengths
of the sequences the edges are labeled with can be taken into account to allow
terminating the search of $v$'s neighborhood before all possible walks of length
2 starting at $v$ have been considered.

@@

\begin{figure}
	\begin{center}
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node[fill=\specialnodecolor] (4) at (3.25, 0) {4};
			\node (5) at (5.0, 0) {5};
			\node (6) at (6.5, 1) {6};
			\node (7) at (6.5, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (4);
			\draw[>->,very thick] (4) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}

@@

		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
			\node (1) at (0.0, -1) {1};
			\node (2) at (0.0, 1) {2};
			\node (3) at (1.5, 0) {3};
			\node (5) at (3.25, 0) {5};
			\node (6) at (4.75, 1) {6};
			\node (7) at (4.75, -1) {7};
			\draw[>->,very thick] (1) edge (3);
			\draw[>->,very thick] (2) edge (3);
			\draw[>->,very thick] (3) edge (5);
			\draw[>->,very thick] (5) edge (6);
			\draw[>->,very thick] (5) edge (7);
		\end{tikzpicture}
		}
	\end{center}
	\caption{Vertex 4 an \KeyTerm{inner vertex} that is part of an unbranched
	path that can be collapsed.  Edge labels are not shown.}
	\label{fig:collapse_unbranched_paths}
\end{figure}

@@

A completely random genome is unlikely to have repeats of significant length
compared to the minimum overlap length.  In this case, transitive reduction is
expected to leave the entire graph as a single unbranched path, provided the
original genome was a single, non-circular string and every base pair of it was
included in at least one read.  Then, collapsing unbranched paths will collapse
the entire bidirected graph into two vertices connected by a single edge.
Finding a walk through the edges of this graph is trivial and will produce an
assembly of the entire genome.  \Figure{fig:random_example} shows the bidirected
string graph that might result from such a genome and data.

If we still assume a genome with no repeats of length greater than the minimum
overlap length, but relax the assumption that the genome is a single,
non-circular string from which every base pair was included in a read, the graph
will instead reduce and collapse to a set of components.  Components
corresponding to non-circular sequence will contain two vertices with an edge
between, while components corresponding to circular sequence will contain one
vertex with an edge connecting it to itself (a loop).  In this case (and as will
be elaborated on later), finding a walk through the bidirected string graph must
be generalized to allow walking each component separately; but again for this
particular case of a random genome, each such walk is still trivial, and the
assembly will be optimal as long as the assembler does not decide to traverse
any of the loops more than one time (this will also be elaborated on more
later), since there are no repeats of substantial length in a hypothetical
random genome.

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}
